Bioconductor workflow for single-cell RNA sequencing: Dimensionality reduction, clustering, and lineage inference

### 2017-06-27 workflow/workflow.Rmd
Edits to text.

scRNA-seq workflows have already been developed, with several useful methods for quality control, visualization, and pre-processing of the data. The workflow for low-level analysis of scRNA-seq data described in [@Lun2016] or the package `scater` [@McCarthy2017] are examples of workflows based on software packages from the open-source Bioconductor Project [@Huber2015] and cover basic steps, including quality control, data exploration, and normalization. In these workflows, single-cell expression data are organized in objects of the SCESet class allowing integrated analysis. However, these workflows are mostly used to prepare the data for further downstream analysis and do not focus on steps like clustering and pseudotime ordering.

Here, we propose an integrated workflow for dowstream analysis, with the following three main steps: dimensionality reduction adjusting for gene and cell-level covariates, robust and stable clustering using resampling-based sequential ensemble clustering, and inference of cell lineages and ordering of the cells by developmental progression. Along the workflow, we use a unique SummarizedExperiment object to create an easy to use sequence of steps, where a different R package is used for each step.

@@

Throughout the workflow, we use an object of class SummarizedExperiment to keep track of the counts and their associated metadata within a single object. The cell-level metadata contain quality control (QC) measures, sequencing batche ID, and cluster labels and lineage labels from the original publication [@Fletcher2017]. Cells with a cluster label of `-2` were not assigned to a cluster in the orginal publication.

@@

Using the Bioconductor R package scone [SD: scone not yet released on Bioconductor], we remove low-quality cells that did not pass our quality control filter, that is ... See Figure \@ref(fig:scone). Davide, would you like to describe your choices for scone here?

@@

Metadata for the cells are stored in the object colData from the SummarizedExperiment object. Cells were processed in 18 different batches.

@@

```{r original}
clus.labels <- colData(core)[, "clusterLabels"]
col_clus <- c("transparent", brewer.pal(12, "Set3"), brewer.pal(8, "Set2"))
col_clus <- col_clus[1:length(unique(clus.labels))]
names(col_clus) <- sort(unique(clus.labels))
table(clus.labels)
```
Note that there is partial confounding of batches and clusters, which could be problematic when correcting for batch effects in the dimensionality reduction step.

@@

In scRNA-seq analysis, dimensionality reduction is often used as a preliminary step prior to downstream analysis where clustering, pseudotime ordering, or differential expression analysis are performed. It allows the data to become more tractable, both from a statistical (cf. curse of dimensionality) and computational point of view. Additionally, technical noise can be reduced while preserving the often intrinsically low dimensional signal of interest [@Peer2017].

Here, we perform dimensionality reduction using bioconductor package zinbwave which allows to fit a zero-inflated negative binomial model (ZINB-WaVE) to get a low-dimensional representation of the data while accounting for zero inflation (dropouts), over-dispersion, and the count nature of the data. The model can include a sample-level intercept, which serves as a global-scaling normalization factor. The user can also include both gene-level and sample-level covariates. The inclusion of observed and unobserved sample-level covariates enables normalization for complex, non-linear effects (often referred to as batch effects), while gene-level covariates may be used to adjust for sequence composition effects, such as gene length and GC-content effects. You can read more details about the ZINB-WaVE model in [@Risso2017].

As with most of the dimensionality reduction methods, the user needs to specify the number of dimensions of the new low dimensional space. Here, we use `K = 50` dimensions in the low dimensional space and adjusted for the batches.

@@

### Normalized values

>>>>

scRNA-seq workflows have already been developed, with several useful methods for quality control, visualization, and pre-processing of the data. The workflow for low-level analysis of scRNA-seq data described in [@Lun2016] or the package `scater` [@McCarthy2017] are examples of workflows based on software packages from the open-source Bioconductor Project [@Huber2015] and cover basic steps, including quality control, data exploration, and normalization. In these workflows, single-cell expression data are organized in objects of the `SCESet` class allowing integrated analysis. However, these workflows are mostly used to prepare the data for further downstream analysis and do not focus on steps like clustering and pseudotime ordering.

Here, we propose an integrated workflow for dowstream analysis, with the following three main steps: dimensionality reduction accounting for zero inflation and adjusting for gene and cell-level covariates, robust and stable clustering using resampling-based sequential ensemble clustering, and inference of cell lineages and ordering of the cells by developmental progression. Along the workflow, we use a unique `SummarizedExperiment` object to create an easy to use sequence of steps, where a different R package is used for each step.

@@

Throughout the workflow, we use an object of class `SummarizedExperiment` to keep track of the counts and their associated metadata within a single object. The cell-level metadata contain quality control (QC) measures, sequencing batche ID, and cluster labels and lineage labels from the original publication [@Fletcher2017]. Cells with a cluster label of `-2` were not assigned to a cluster in the orginal publication.

@@

Using the Bioconductor R package `scone` [SD: scone not yet released on Bioconductor], we remove low-quality cells that did not pass our quality control filter, that is ... See Figure \@ref(fig:scone). Davide, would you like to describe your choices for scone here?

@@

Metadata for the cells are stored in the slot `colData` from the `SummarizedExperiment` object. Cells were processed in 18 different batches.

@@

```{r original}
clus.labels <- colData(core)[, "clusterLabels"]
col_clus <- c("transparent", brewer.pal(12, "Set3"), brewer.pal(8, "Set2"))
col_clus <- col_clus[1:length(unique(clus.labels))]
names(col_clus) <- sort(unique(clus.labels))
table(clus.labels)
```
Note that there is partial confounding [SD: Did you mean nesting?] of batches and clusters, which could be problematic when correcting for batch effects in the dimensionality reduction step.

@@

In scRNA-seq analysis, dimensionality reduction is often used as a preliminary step prior to downstream analyses, such as clustering, cell lineage and pseudotime ordering, and differential expression (DE) analysis. This allows the data to become more tractable, both from a statistical (cf. curse of dimensionality) and computational point of view. Additionally, technical noise can be reduced while preserving the often intrinsically low dimensional signal of interest [@Peer2017] [@Risso2017 ][SD: Ref to ZIFA.].

Here, we perform dimensionality reduction using the Bioconductor R package `zinbwave`, which fits a zero-inflated negative binomial model (ZINB-WaVE) that accounts for zero inflation (dropouts), over-dispersion, and the count nature of the data. The model can include a sample-level intercept, which serves as a global-scaling normalization factor. The user can also include both gene-level and sample-level covariates. The inclusion of observed and unobserved sample-level covariates enables normalization for complex, non-linear effects (often referred to as batch effects), while gene-level covariates may be used to adjust for sequence composition effects, such as gene length and GC-content effects. For greated detail about the ZINB-WaVE model and estimation procedure, please refer to the original manuscript [@Risso2017].

[SD: Include Fig. 1 from [@Risso2017]?]
As with most dimensionality reduction methods, the user needs to specify the number of dimensions for the new low-dimensional space. Here, we use `K = 50` dimensions and adjust for the batch effects.

@@

### Normalization
